# Statistical distributions
## Normal Distribution

```{r, eval=TRUE}
x <- seq(-6, 6, length=100)
y1 <- dnorm(x, sd=1)
y2 <- dnorm(x, sd=0.5)
y3 <- dnorm(x, sd=2)
plot(x, y1, 
     xlim=c(-6,6), ylim=c(0,0.8),
     type="l", lwd=2,
     xlab="x value", ylab="Density",
     main="Normal distribution")
lines(x, y2, col="red")
lines(x, y3, col="green")
legend("topright", legend=c("sd = 1", "sd = 0.5", "sd = 2"),
       col=c("black", "red", "green"), 
       lty=c(1,1,1), lwd=2)

# Area under the curve
# For normal distribution auc = 1 (probability for all)
library(DescTools)
AUC(x, y1)
AUC(x,y2)
AUC(x,y3) # to broad distribution, some samples out of the area
```

## Bernoulli Distribution

## Binomial Distribution

## Geometric Distribution

## Uniform Distributions

## Poisson Distribution
**Poisson distribution** is a discrete probability distribution that expresses the probability of a given number of events occurring in a 
fixed interval of time or space if these events occur with a known constant mean rate and independently of the time since the last event. [wiki](https://en.wikipedia.org/wiki/Poisson_distribution)
Also knows as **Poisson Process**.  

$$P(k) = e^{\lambda}\frac{\lambda^k}{k!}$$

The Poisson distribution is also the limit of a binomial distribution, for which the probability of success for each trial equals λ divided by the number of trials, as the number of trials approaches infinity.  

According to the Central Limit Theorem for big $\lambda$ the Normal distribution will be the approximation for the Poisson distribution.  

Criteria of the Poisson process:  
1. Events are independent of each other. The occurrence of one event does not affect the probability another event will occur.
2. The average rate (events per time period) is constant.
3. Two events cannot occur at the same time.

Models:  
* Number of calls per minute in a call center.  
* Number of decay events that occur from a radioactive source in a given observation period.  
* The number of meteorites greater than 1 meter diameter that strike Earth in a year.  
* The number of patients arriving in an emergency room between 10 and 11 pm.  
* The number of laser photons hitting a detector in a particular time interval.

Example:  
12 cars cross the bridge in 1 min in average. What is the probability that 17 or more cars will cross the bridge?  
```{r}
ppois(16, lambda=12, lower=FALSE)
```

## Exponential Distribution
**Exponential distribution** is the probability distribution of the time between events in a Poisson point process, i.e., 
a process in which events occur continuously and independently at a constant average rate. [wiki](https://en.wikipedia.org/wiki/Exponential_distribution)

$$P(x) = \lambda e ^{ - \lambda x }$$

It is a particular case of the gamma distribution.

Suppose the mean checkout time of a supermarket cashier is 3 minutes. Find the probability of a customer checkout being completed by the cashier in less than 2 minutes.  

```{r}
pexp(2, rate=1/3)
```

## Chi-squared Distribution
In probability theory and statistics, the $\{chi}^2$ distribution with k degrees of freedom is the distribution of a sum of the squares of 
$k$ independent standard normal random variables. [wiki](https://en.wikipedia.org/wiki/Chi-square_distribution)
If $X_1,X_2,…,X_m$ are $m$ independent random variables having the standard normal distribution, then the following quantity follows a **Chi-Squared distribution** with $m$ degrees of freedom. Its mean is $m$, and its variance is $2m$.

$$V = X_1^2 + X_2^2 + ... + X_m^{2} \tilde{} X_{(m)}^2$$

Summ of squares of standard normal, independen random variables are distributed according to the chi-square distribution with k degrees of freedom, where k is the number of random variables being summed.  

Chi-squared distribution applications:  
* Chi-square test of independence in contingency tables  
* Chi-square test of goodness of fit of observed data to hypothetical distributions  
* Likelihood-ratio test for nested models  
* Log-rank test in survival analysis  
* Cochran–Mantel–Haenszel test for stratified contingency tables  
```{r}
x = seq(0,25, length=100)
v = dchisq(x, df=7)
plot(x,v, type='l')
# quantiles of chi-squared
q = c(.25, .5, .75, .999)
qchi = qchisq(q, df = 7, lower.tail = TRUE)
abline(v=qchi, lty=2, col='red')
qchi
```

**Sources**  
[Seven Must-Know Statistical Distributions and Their Simulations for Data Science](https://towardsdatascience.com/seven-must-know-statistical-distributions-and-their-simulations-for-data-science-681c5ac41e32#:~:text=A%20statistical%20distribution%20is%20a,the%20random%20value%20it%20models.)
[Probability distributions](http://www.r-tutor.com/elementary-statistics/probability-distributions)
Hartmann, K., Krois, J., Waske, B. (2018): E-Learning Project SOGA: Statistics and Geospatial Data Analysis. Department of Earth Sciences, Freie Universitaet Berlin.  
