# Chi-squared test


## Multinomial Goodness of Fit
A population is called multinomial if its data is categorical and belongs to a collection of discrete non-overlapping classes.

The null hypothesis for goodness of fit test for multinomial distribution is that the observed frequency fi is equal to an expected count $$e_i$$ in each category. 
It is to be rejected if the p-value of the following **Chi-squared test** statistics is less than a given significance level α.

Example
Survey response about the student’s smoking habit: "Heavy", "Regul" (regularly), "Occas" (occasionally) and "Never". 
The Smoke data is multinomial.

```{r}
library(MASS)
levels(survey$Smoke) 
smoke_freq = table(survey$Smoke) 
smoke_freq

# estimated probabilities
smoke_prob = c(heavy = .045, 
               never = .795, 
               occas = .085, 
               regul = .075)
```

Determine whether the sample data in *smoke_freq* supports estimated probabilities in *smoke_prob* at .05 significance level.

```{r}
chisq.test(smoke_freq, p=smoke_prob)
```

As the p-value 0.991 is greater than the .05 significance level, we do not reject the null hypothesis that the sample data in survey supports the smoking statistics.  

**Sources**

[Multinomial Goodness of Fit](https://www.r-tutor.com/elementary-statistics/goodness-fit/multinomial-goodness-fit)
