# Analysis of Variance (ANOVA)
## One-way ANOVA
**variance** = SS/df, where SS - sum of squares and df - degree of freedom  
\(SS = \displaystyle\sum_{i=1}^{n}{(x_i - \mu)^2}\), where  
\(\mu\) is the sample **mean**  
**n** is the sample **size** 

\(var(x) = \frac{1}{n}{\displaystyle\sum_{i=1}^{n}{(x_i - \mu)^2}}\)  



SST = SSE + SSC = W + B, where  
SST - Total Sum of Squares  
SSE - Error Sum of Squares - within (W)  
SSC - Sum of Squares Columns (treatmens) - between (B)

C - columns (treatments)  
N - total number of observations  

Mean squared of columns - MSC = SSC/df_columns, where df_columns = C-1  
Mean squared of error - MSE = SSE/df_error, where df_error = N-C  
Sum of squares (total) - SST, where df_total = N-1
F-statistics - F = MSC/MSE

Let's calculate degree of freedom for our example:  
df_columns = 3-1 = 2, MSC = SSC/2  
df_error = 21-3 = 18, MSE = SSE/18  
df_total = 21-1 = 20

```{r}
# 3 groups of students with scores (1-100):
a = c(82,93,61,74,69,70,53)
b = c(71,62,85,94,78,66,71)
c = c(64,73,87,91,56,78,87)

sq = function(x) { sum((x - mean(x))^2) }

sq(a)
sq(b)
sq(c)

```

Using R packages:
```{r}
# data
# Number of calories consumed by month:
may <- c(2166, 1568, 2233, 1882, 2019)
sep <- c(2279, 2075, 2131, 2009, 1793)
dec <- c(2226, 2154, 2583, 2010, 2190)

d <- stack(list(may=may, sep=sep, dec=dec))
d
names(d)

oneway.test(values ~ ind, data=d, var.equal=TRUE)
```

```{r}
# alternative using aov
res <- aov(values ~ ind, data = d)
res
summary(res)
```

## Sources
Example for one-way ANOVA: [youtube](https://www.youtube.com/playlist?list=PLIeGtxpvyG-KA-BLkL391X__r0kU4_hm5) by Brandon Foltz  
