# Split data into train and test subsets
  
Here you can find several simple approaches to split data into train and test subset to fit and to test parameters of your model. 
We want to take 0.8 of our initial data to train our model.  
Data: `datasets::iris`.

1. First approach is to create a **vector containing randomly selected row ids** and to apply this vector to split data.  

```{r}
inTrain = sample(nrow(iris), nrow(iris)*0.8)

# split data
train = iris[inTrain, ]
test = iris[-inTrain, ]
```

2. The same idea to split data as before using `caret` package.  
The advantage is that `createDataPartition` function allows to split data many `times` and  use these subsets to estimate parameters of our model.  
```{r, message=F}
library(caret)
trainIndex <- createDataPartition(iris$Species, p=.8,
                                  list = FALSE,        # if FALSE - create a vector/matrix, if TRUE - create a list
                                  times = 1)           # how many subsets
# split data
train <- iris[trainIndex, ]
test <- iris[-trainIndex, ]
```

3. Another approch is to create a **logical vecotor** containing randomly distributed true/false and apply this vector to subset data.  
```{r}
inTrain = sample(c(TRUE, FALSE), nrow(iris), replace = T, prob = c(0.8,0.2))

# select data
train = iris[inTrain, ]
test = iris[!inTrain, ]
```

4. Using [`caTools`](https://cran.r-project.org/web/packages/caTools/index.html).
```{r}
library(caTools)
inTrain = sample.split(iris, SplitRatio = .8)
train = subset(iris, inTrain == TRUE)
test  = subset(iris, inTrain == FALSE)
```

5. Using [`dplyr`](https://cran.r-project.org/web/packages/dplyr/)
```{r, message=F}
library(dplyr)
iris$id <- 1:nrow(iris)
train <- iris %>% dplyr::sample_frac(.8)
test  <- dplyr::anti_join(iris, train, by = 'id')
```
